Syntax and Typing for Core Cedille

### 2018-02-14 docs/cedille-core/spec.tex
some tweaks
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@29278 bd178f5f-7bea-4014-844f-495ca19b8805

\section{Syntax}

@@

\begin{figure}
\[
\begin{array}{llll}
  \textit{variables}\ x & \ & \ &\ \\
  \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a term variable} \\
  \ &\ & \star&\textnormal{kind for types}\\
  \ &\ & \Box&\textnormal{sole superkind} \\
  \ &\ & t.1 & \textnormal{project first view of a dependent intersection}\\
  \ &\ & t.2 & \textnormal{project second view of a dependent intersection}\\
  \ &\ & \beta\{t\} &\textnormal{proof of a true equation, where the proof erases to }t  \\
  \ &\ & \varsigma\ t &\textnormal{symmetry of equality}  \\
  \ &\ & t\ t' & \textnormal{application of term to term}\\
  \ &\ & t\ \mhyph t' & \textnormal{application of a term to an erased argument}\\
  \ &\ & \rho\ t\ @\ x.t'\ \mhyph\ t''&\textnormal{equality elimination by type-guided rewriting}  \\ 
  \ &\ & \abs{\forall}{x}{t}{t'}\ &\textnormal{implicit product (quantify over erased argument)} \\
  \ &\ & \abs{\Pi}{x}{t}{t'}\ &\textnormal{explicit product (usual }\Pi\textnormal{-type)} \\
  \ &\ & \abs{\iota}{x}{T}{T'}\ &\textnormal{dependent intersection} \\
  \ &\ & \abs{\lambda}{x}{t}{t'} & \textnormal{usual }\lambda\textnormal{-abstraction} \\
  \ &\ & \abs{\Lambda}{x}{t}{t'} & \textnormal{erased }\lambda\textnormal{-abstraction}\\
  \ &\ & [ t , t' \ @\ x.t'' ] & \textnormal{introduce dependent intersection}\\
  \ &\ & \phi\ t\ \mhyph\ t'\ \{t''\} & \textnormal{when }t\textnormal{ proves }t'\textnormal{ and }t''
                                   \textnormal{ are equal, erase to }t'' \\
  \ & \ & [ x = t : t' ]\ \mhyph\ t'' & \textnormal{let }x\textnormal{ equal }t\textnormal{ of type }t'\textnormal{ in }t'' \\
  \ &\ & \{ t \simeq t' \} &\textnormal{equality between terms}
\end{array}
  \]
  \caption{Sytax for core Cedille}
  \label{fig:syntax}
\end{figure}

@@

\begin{itemize}
\item The rules are applied bottom-up, and a judgment $\Gamma\vdash t : t'$ represents a call to compute
  a type $t'$ given a context $\Gamma$ and a term $t$.
\item Some of the premises of the rules state that a computed type should have a certain form (e.g., be a $\Pi$-type).  It may happen, though, that the computed type $\beta$-reduces to a $\Pi$-type but is not literally one.  To handle this case, one should head-normalize the computed types for premises that require a specific form.
\item Where two premises use the same meta-variable, one must check definitional equality of the terms in question (for example, the domain-part of a $\Pi$-type and the type of an argument, in the application typing rule).
  The definitional equality relation, which we can denote $\Gamma \vdash t =_{\beta\eta} t'$,
  is the usual relation of $\beta\eta$-equivalence of
terms in pure untyped lambda calculus, extended congruentially for the
typing constructs $\forall$, $\Pi$, $\iota$, and $\simeq$, and also
extended to make use of let-definitions $x = t : t'$ contained in
$\Gamma$ (by replacing $x$ with $t$ when checking definitional
equality).  This same mechanism can be used for global definitions, as
well.
\end{itemize}

\begin{figure}
  \[
  \begin{array}{lll}
    \infer{\Gamma\vdash \star : \Box}{\ } &
    \infer{\Gamma\vdash\abs{\Pi}{x}{t}{t'} : s}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : s} &
    \infer{\Gamma\vdash\abs{\forall}{x}{t}{t'} : \star}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : \star} 
    \\ \\
    \infer{\Gamma, x : t \vdash x : t}{\ } &
    \infer{\Gamma\vdash t\ t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\Pi}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash\abs{\iota}{x}{t}{t'} : \star}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : \star} 
    \\ \\
    \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash\abs{\lambda}{x}{t}{t'} : \abs{\Pi}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\Pi}{x}{t}{t''} : s}  &
    \infer{\Gamma\vdash\abs{\Lambda}{x}{t}{t'} : \abs{\forall}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\forall}{x}{t}{t''} : s} \\ \\
    \infer{\Gamma\vdash t.1 : t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &
    \infer{\Gamma\vdash t.2 : [t/x]t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2}
    \\ \\
    \infer{\Gamma\vdash t \simeq t' : \star}{\textit{FV}(t\ t')\subseteq \textit{dom}(\Gamma) } &
    \infer{\Gamma\vdash \beta\{t\} : \{ t' \simeq t'\}}{\ } &
    \infer{\Gamma\vdash \rho\ t\ @\ x.t'\ \mhyph\ t'' : [t_2/x]t'}{\Gamma \vdash t : \{ t_1 \simeq t_2 \} \quad \Gamma\vdash t'' : [t_1/x]t'}
    \\ \\
    \infer{\Gamma\vdash \varsigma\ t : \{ t_2 \simeq t_1\}}{\Gamma\vdash t : \{t_1 \simeq t_2\}} &
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'} &
    \infer{\Gamma\vdash [ x = t_1 : t' ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma, x = t_1 : t' \vdash t_2 : t''} 
  \end{array}
  \]
  \caption{Type-checking rules for Core Cedille}
  \label{fig:rules}
\end{figure}

@@

>>>>

\section{Syntax}
\label{sec:syntax}

@@

\begin{figure}
\[
\begin{array}{llll}
  \textit{term variables}\ u & \ & \ &\ \\
  \textit{type variables}\ X & \ & \ &\ \\
  \textit{kind variables}\ k & \ & \ &\ \\
  \textit{variables}\ x & ::= & u\ |\ X\ |\ k &\ \\  
  \textit{pure terms}\ p & ::= & u\ |\ p\ p'\ |\ \absu{\lambda}{u}{p} &\ \\
  \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a term variable} \\
  \ &\ & \star&\textnormal{kind for types}\\
  \ &\ & \Box&\textnormal{sole superkind} \\
  \ &\ & t.1 & \textnormal{project first view of a dependent intersection}\\
  \ &\ & t.2 & \textnormal{project second view of a dependent intersection}\\
  \ &\ & \beta\{t\} &\textnormal{proof of a true equation, where the proof erases to }t  \\
  \ &\ & \varsigma\ t &\textnormal{symmetry of equality}  \\
  \ &\ & t\ t' & \textnormal{application of term to term}\\
  \ &\ & t\ \mhyph t' & \textnormal{application of a term to an erased argument}\\
  \ &\ & \rho\ t\ @\ x.t'\ \mhyph\ t''&\textnormal{equality elimination by type-guided rewriting}  \\ 
  \ &\ & \abs{\forall}{x}{t}{t'}\ &\textnormal{implicit product (quantify over erased argument)} \\
  \ &\ & \abs{\Pi}{x}{t}{t'}\ &\textnormal{explicit product (usual }\Pi\textnormal{-type)} \\
  \ &\ & \abs{\iota}{x}{T}{T'}\ &\textnormal{dependent intersection} \\
  \ &\ & \abs{\lambda}{x}{t}{t'} & \textnormal{usual }\lambda\textnormal{-abstraction} \\
  \ &\ & \abs{\Lambda}{x}{t}{t'} & \textnormal{erased }\lambda\textnormal{-abstraction}\\
  \ &\ & [ t , t' \ @\ x.t'' ] & \textnormal{introduce dependent intersection}\\
  \ &\ & \phi\ t\ \mhyph\ t'\ \{t''\} & \textnormal{when }t\textnormal{ proves }t'\textnormal{ and }t''
                                   \textnormal{ are equal, erase to }t'' \\
  \ & \ & [ x = t : t' ]\ \mhyph\ t'' & \textnormal{let }x\textnormal{ equal }t\textnormal{ of type }t'\textnormal{ in }t'' \\
  \ &\ & \{ p \simeq p' \} &\textnormal{equality between pure terms}
\end{array}
  \]
  \caption{Sytax for core Cedille}
  \label{fig:syntax}
\end{figure}

\subsection{Syntactic check for being a term}
\label{sec:termcheck}

It is important that we only form equations between terms, because
Cedille's semantics does not support forming equations except where
both sides of the equation are terms.  Terms can be syntactically
distinguished from types and kinds as long as we use different
syntactic categories for term variables, type variables, and kind
variables.  That is why the syntax of Figure~\ref{fig:syntax}
distinguishes these.  The typing rule below for forming equations
requires the sides to be pure (i.e., unannotated) terms.

@@

\begin{itemize}
\item The rules are applied bottom-up, and a judgment $\Gamma\vdash t : t'$ represents a call to compute
  a type $t'$ given a context $\Gamma$ and a term $t$.
\item Some of the premises of the rules state that a computed type should have a certain form (e.g., be a $\Pi$-type).  It may happen, though, that the computed type $\beta$-reduces to a $\Pi$-type but is not literally one.  To handle this case, one should head-normalize the computed types for premises that require a specific form.
\item Where two premises use the same meta-variable, one must check definitional equality of the terms in question (for example, the domain-part of a $\Pi$-type and the type of an argument, in the application typing rule).
  The definitional equality relation, which we can denote $\Gamma \vdash t =_{\beta\eta} t'$,
  is the usual relation of $\beta\eta$-equivalence of
terms in pure untyped lambda calculus, extended congruentially for the
typing constructs $\forall$, $\Pi$, $\iota$, and $\simeq$, and also
extended to make use of let-definitions $x = t : t'$ contained in
$\Gamma$ (by replacing $x$ with $t$ when checking definitional
equality).  This same mechanism can be used for global definitions, as
well.
\item Note that the formation rules for $\forall$- and $\Pi$-abstractions use
  a premise $\textit{Var}(x,s)$ to check
  that the variable $x$ is a legal form of variable given the kind $s$.  This
  judgment is defined by these rules:
  \[
  \begin{array}{lll}
    \infer{\textit{Var}(u,\star)}{\ } & \ \ \ & \infer{\textit{Var}(X,\Box)}{\ }
  \end{array}
  \]

\end{itemize}

\begin{figure}
  \[
  \begin{array}{lll}
    \infer{\Gamma\vdash \star : \Box}{\ } &
    \infer{\Gamma\vdash\abs{\Pi}{x}{t}{t'} : s'}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : s' \quad \textit{Var}(x,s)} &
    \infer{\Gamma\vdash\abs{\forall}{x}{t}{t'} : \star}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : \star \quad \textit{Var}(x,s)} 
    \\ \\
    \infer{\Gamma, x : t \vdash x : t}{\ } &
    \infer{\Gamma\vdash t\ t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\Pi}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash\abs{\iota}{u}{t}{t'} : \star}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : \star} 
    \\ \\
    \infer{\Gamma\vdash t.2 : [t/x]t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
    \infer{\Gamma\vdash\abs{\lambda}{x}{t}{t'} : \abs{\Pi}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\Pi}{x}{t}{t''} : s}  &
    \infer{\Gamma\vdash\abs{\Lambda}{x}{t}{t'} : \abs{\forall}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\forall}{x}{t}{t''} : s} \\ \\
    \infer{\Gamma\vdash t.1 : t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &
    \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2 \quad \Gamma \vdash \abs{\iota}{x}{t_1}{t_2} : \star}
    \\ \\
    \infer{\Gamma\vdash \beta\{t\} : \{ t' \simeq t'\}}{\Gamma\vdash \{ t' \simeq t' \} : \star } &
    \infer{\Gamma\vdash \{ p \simeq p'\} : \star}{\textit{FV}(p\ p')\subseteq \textit{dom}(\Gamma)} &
    \infer{\Gamma\vdash \rho\ t\ @\ x.t'\ \mhyph\ t'' : [t_2/x]t'}{\Gamma \vdash t : \{ t_1 \simeq t_2 \} \quad \Gamma\vdash t'' : [t_1/x]t'}
    \\ \\
    \infer{\Gamma\vdash \varsigma\ t : \{ t_2 \simeq t_1\}}{\Gamma\vdash t : \{t_1 \simeq t_2\}} &
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'} &
    \infer{\Gamma\vdash [ k = t_1 : \Box ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : \Box \quad
                                                               \Gamma, k = t_1 : t' \vdash t_2 : t''} 
    \\ \\
    \multicolumn{2}{l}{
    \infer{\Gamma\vdash [ x = t_1 : t' ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma \vdash t' : s \quad \textit{Var}(x,s) \quad
      \Gamma, x = t_1 : t' \vdash t_2 : t''} } &
    \     
  \end{array}
  \]
  \caption{Type-checking rules for Core Cedille}
  \label{fig:rules}
\end{figure}

@@

\appendix 
\section{Change log}

\begin{itemize}
\item Feb. 14, 2018:
  \begin{itemize}
  \item added another premise to the rule for introducing dependent intersections, to
    make sure that the dependent intersection itself if typable.
  \item Equations are now required to be only between pure
    (unannotated) terms.  This is a syntactic check, which is now
    described in Section~\ref{sec:termcheck}.  It requires different
    lexical classes for term, type, and kind variables.  In turn, this
    requires that the typing rules check, with a new helper judgment
    $\textit{Var}(x,s)$, that we are using the proper class of
    variable when we introduce a local variable.

  \end{itemize}
\end{itemize}
